---
title: "Advanced Data Wrangling in R"
author: "Kelsey E. Gonzalez"
date: 2021-05-25
summary: "A 9 hour advanced data wrangling workshop conducted with the Critical Path Institute in Tucson Arizona "
tags: ["Workshop"]
image:
  focal_point: ''
  preview_only: no
projects: []
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      results = FALSE,
                      message = FALSE,
                      warning = FALSE)
```

# Welcome

This is the course site for the Advanced Data Wrangling in R workshop created by Kelsey Gonzalez for the Critical Path Institute in Tucson Arizona from May 26-28, 2021.

This nine-hour hands-on workshop is an exploration of the Tidyverse as a tool to wrangle your data in preparation for analysis. 

Day 1 will cover basic dplyr funcitonality including `select`, `filter`, `arrange`, `mutate`, `group_by` and `summarize` and then moves into the concept of tidy data and pivots. 

<div class="shareagain" style="min-width:300px;margin:1em auto;">
  <iframe src="https://kelseygonzalez.github.io/cpath-training/day-01-basics.html" width="500" height="300" style="border:2px solid currentColor;" loading="lazy" allowfullscreen></iframe>
  <script>fitvids('.shareagain', {players: 'iframe'});</script>
</div>

Day 2 will cover special variable types and advanced creation of new variables. Specifically, we will cover `case_when`, factors with `forcats`, dates with `lubridate`, and strings and regular expressions with `stringr`. 

<div class="shareagain" style="min-width:300px;margin:1em auto;">
  <iframe src="https://kelseygonzalez.github.io/cpath-training/day-02-variables.html" width="500" height="300" style="border:2px solid currentColor;" loading="lazy" allowfullscreen></iframe>
  <script>fitvids('.shareagain', {players: 'iframe'});</script>
</div>

Day 3 will cover relational data and joins in addition to more advanced and cutting edge dplyr functions like `across`, `colwise`, `rowwise`. After diving into `purrr` for iteration, we will dive into case studies. 

# Prework

We will use Rstudio desktop for this workshop. You're also welcome to utilize Rstudio cloud.
- Install [R](https://cloud.r-project.org/) and [RStudio Desktop](https://rstudio.com/products/rstudio/download/) on your computer. 

You can find step-by-step instructions for installing these here: [macOS](https://www.youtube.com/watch?v=GM88tYlEy_g), [Windows](https://www.youtube.com/watch?v=JRKmZK5-6aE).

- Install the following packages:

```{r, eval=FALSE, echo=TRUE}
# From CRAN
install.packages("tidyverse")
install.packages("NHANES")
install.packages("janitor")
install.packages("gapminder")

```

# Links

- Link to this website: [https://bit.ly/cpath-wrangling](https://bit.ly/cpath-wrangling)
- Day 1: [Slides](cpath-training/day-01-basics.html)
- Day 2: [Slides](cpath-training/day-02-variables.html)
- Day 3: [Slides](cpath-training/day-03-advanced.html)